Chord: A Versatile Platform for Program Analysis

### 2011-05-19 main/doc/predefined.tex
updated documentation

For instance, the following command runs a basic may-alias and call-graph
analysis (called 0CFA) provided in Chord:

@@

Chord offers several choices for computing points-to and call-graph information
of Java programs.  In each of these choices, the points-to and call-graph
information is computed simultaneously (called ``on-the-fly call-graph
construction" in the literature in contrast to ``ahead-of-time call-graph
construction" in which the call graph is computed first followed by points-to
information).  On-the-fly approaches are more precise because, in a dynamically
dispatching language like Java, as more points-to facts are discovered, more
(dynamically dispatched) methods are deemed reachable, thereby growing the call
graph; the code in these newly added methods in turn results in more points-to
facts.

@@

The context-insensitive points-to/call-graph analysis treats each object
allocation site as a separate abstract memory location; in other words, it can
distinguish objects created at different sites but not those created at the same
site.  Additionally, it is field-sensitive, in that it distinguishes between
different instance fields of the same object, but array-insensitive, in that it
cannot distinguish between different elements of the same array; all array
elements are modeled using a distinguished hypothetical instance field (that has
index 0 in domain F).

@@

\begin{itemize}
\item {\it Call-graph information:}
\begin{itemize}
\item
rootM subset M contains the set of entry methods that may be reachable; this
includes the program's main method as well as each static initializer method
that may be reachable from the main method.
\item
reachableM subset M contains the set of methods that may be reachable from the
program's main method.
\item
IM subset (I $\times$ M) contains tuples \texttt{(i,m)} such that call site
\texttt{i} may call method \texttt{m}.
%\item
% MM subset (M $\times$ M) contains tuples \texttt{(m1,m2)} such that
%method \texttt{m1} may call method \texttt{m2}.
%\item
%reachableI subset I contains the set of call sites that may be reachable.
%\item
%reachableT subset T contains the set of classes that may be reachable.
\end{itemize}
\item
{\it Points-to information:}
\begin{itemize}
\item
FH subset (F $\times$ H) contains tuples \texttt{(f,h)} such that static field
(i.e.  global variable) \texttt{f} may point to an object allocated at site
\texttt{h}.
\item
VH subset (V $\times$ H) contains tuples \texttt{(v,h)} such that local variable
\texttt{v} may point to an object allocated at site \texttt{h}.
\item
HFH subset (H $\times$ F $\times$ H) contains tuples \texttt{(h1,f,h2)} such
that instance field \texttt{f} of some object allocated at site \texttt{h1} may
point to some object allocated at site \texttt{h2}.
\end{itemize}
\end{itemize}

@@

In a context-sensitive analysis, there is no longer just one abstract memory
location per object allocation site.  Rather, the set of objects a reference can
point to depends on the \textit{context} in which the method containing the
reference is called.  Whereas a context-insensitive analysis talks about the
domain of methods (M) and the domain of allocation sites (H), a
context-sensitive analysis talks about the domain of abstract contexts, labeled
C.  Elements of domain C contain both abstract calling contexts and abstract
objects. (These are merged for reasons described below.)

Chord has several context-sensitive analyses, but they all expose the same
relations, which are described below:

\begin{itemize}
\item
{\it Context information:}
\begin{itemize}
\item
C is the domain of all abstract calling contexts and abstract objects.  Each
element in this domain is a sequence of zero or more sites, where each site can
be a call site or an object allocation site.  A sequence may have mixed call
sites and object allocation sites.  The most significant site (i.e. the first
site) in each sequence is called the head; the remaining sub-sequence is called
the tail.  The below three relations relate a sequence with its head and tail.
\item
CC subset (C $\times$ C) contains tuples \texttt{(c1,c2)} such that context
\texttt{c2} is the tail of context \texttt{c1}.
\item
CH subset (C $\times$ H) contains tuples \texttt{(c,h)} such that object
allocation site \texttt{h} is the head of context \texttt{c}.
\item
CI subset (C $\times$ I) contains tuples \texttt{(c,i)} such that call site
\texttt{i} is the head of context \texttt{c}.
\end{itemize}
\item
{\it Call-graph information:}
\begin{itemize}
\item
rootCM subset (C $\times$ M) contains tuples \texttt{(c,m)} such that method
\texttt{m} is an entry method in context \texttt{c}.
\item
reachableCM subset (C $\times$ M) contains tuples \texttt{(c,m)} such that
method \texttt{m} may be reachable in context \texttt{c}.
\item
CICM subset (C $\times$ I $\times$ C $\times$ M) contains tuples
\texttt{(c1,i,c2,m)} such that call site \texttt{i} in context \texttt{c1} may
call method \texttt{m2} in context \texttt{c2}.
\end{itemize}
\item
{\it Points-to information:}
\begin{itemize}
\item
FC subset (F $\times$ C) contains tuples \texttt{(f,o)} such that static field
(i.e. global variable) \texttt{f} may point to object \texttt{o}.
\item
CVC subset (C $\times$ V $\times$ C) contains tuples \texttt{(c,v,o)} such that
local variable \texttt{v} may point to object \texttt{o} in context \texttt{c}
of that variable's declaring method.  Note that both \texttt{o} and \texttt{c}
are elements of domain C.
\item
CFC subset (C $\times$ F $\times$ C) contains tuples \texttt{(o1,f,o2)} such
that instance field \texttt{f} of object \texttt{o1} may point to object
\texttt{o2}.  \end{itemize}
\end{itemize}

{\bf Under Construction: explain object-sensitive vs context-sensitive, plus how
to invoke.}

>>>>

For instance, the following command runs a basic points-to and call-graph
analysis (called 0CFA) provided in Chord:

@@

Chord provides several analyses with different precision/scalability tradeoff for
computing points-to and call-graph information of Java programs.  All these
analyses are defined under directory \code{src/chord/analyses/alias/}.

In all these analyses, the points-to and call-graph information is computed
simultaneously (called ``on-the-fly call-graph construction" in the literature
in contrast to ``ahead-of-time call-graph construction" in which the call graph
is computed first followed by points-to information).  On-the-fly approaches are
more precise because, in a dynamically dispatching language like Java, as more
points-to facts are discovered, more (dynamically dispatched) methods are deemed
reachable, thereby growing the call graph; the code in these newly added methods
in turn results in more points-to facts.

Broadly, the various analyses are characterized by two aspects (1) flow
sensitive vs. insensitive and (2) context sensitive vs. insensitive.

@@

The context-insensitive points-to/call-graph analysis derives its name from the
fact that it analyzes each method in a single abstract context.
Additionally, the analysis is:
\begin{itemize}
\item
{\it object-insensitive}: it cannot distinguish between
different objects created at the same object allocation site; the reason is that
it treats each object allocation site as a separate abstract memory location.
\item
{\it field-sensitive}: it distinguishes between
different instance fields of the same object.
\item
{\it array-insensitive}: it cannot distinguish between
different elements of the same array; all array elements are modeled using a
distinguished hypothetical instance field (that has index 0 in domain F).
\end{itemize}

@@

\begin{itemize}
\item {\it Call-graph information:}
\begin{itemize}
\item
rootM subset M contains the set of entry methods that may be reachable; this
includes the program's main method as well as each class initializer method
that may be reachable from the main method.
\item
reachableM subset M contains the set of methods that may be reachable from the
program's main method.
\item
IM subset (I $\times$ M) contains tuples \texttt{(i,m)} such that call site
\texttt{i} may call method \texttt{m}.
%\item
% MM subset (M $\times$ M) contains tuples \texttt{(m1,m2)} such that
%method \texttt{m1} may call method \texttt{m2}.
%\item
%reachableI subset I contains the set of call sites that may be reachable.
%\item
%reachableT subset T contains the set of classes that may be reachable.
\end{itemize}
\item
{\it Points-to information:}
\begin{itemize}
\item
FH subset (F $\times$ H) contains tuples \texttt{(f,h)} such that static field
(i.e.  global variable) \texttt{f} may point to an object allocated at site
\texttt{h}.
\item
VH subset (V $\times$ H) contains tuples \texttt{(v,h)} such that local variable
\texttt{v} may point to an object allocated at site \texttt{h}.
\item
HFH subset (H $\times$ F $\times$ H) contains tuples \texttt{(h1,f,h2)} such
that instance field \texttt{f} of some object allocated at site \texttt{h1} may
point to some object allocated at site \texttt{h2}.
\end{itemize}
\end{itemize}

@@

In a context-sensitive analysis, each method is analyzed in potentially multiple
contexts, in contrast to a context-insensitive analysis, in which each
method is always analyzed in a single context.  Additionally, a
context-sensitive analysis can be {\it object-sensitive} in that it can distinguish
between different objects created at the same object allocation site, whereas
a context-insensitive analysis is always object-insensitive.

Different kinds of context- and object-sensitivity have been proposed in the
literature.  Chord's context-sensitive points-to/call-graph analysis is highly
parametric in the kind of context- and object-sensitivity.  It allows:
\begin{itemize}
\item
each method to be analyzed using a different kind of context sensitivity,
namely, one of context insensitivity, k-CFA, k-object-sensitivity, and
copy-context-sensitivity;
\item
each local variable to be analyzed context sensitively or insensitively; and
\item
a different 'k' value to be used for each object allocation site and method
call site.
\end{itemize}

The analysis can be called multiple times and in each invocation it can
incorporate feedback from a client to adjust the precision of the points-to
information and call graph computed subsequently by the points-to/call-graph
analyses.  Clients can indicate in each invocation:

\begin{itemize}
\item
Which methods must be analyzed context sensitively (in addition to those already being
analyzed context sensitively in the previous invocation of this analysis) and using
what kind of context sensitivity; the remaining methods will be analyzed context
insensitively (that is, in the lone 'epsilon' context).
\item
Which local variables of reference type must be analyzed context sensitively (in
addition to those already being analyzed context sensitively in the previous
invocation of this analysis); the remaining ones will be analyzed context insensitively
(that is, their points-to information will be tracked in the lone 'epsilon' context).
\item
The object allocation sites and method call sites whose 'k' values must be incremented
(over those used in the previous invocation of this analysis).
\end{itemize}

\begin{framed}
\begin{verbatim}
ant -Dchord.work.dir=<WORK_DIR> -Dchord.run.analyses=cspa-hybrid-dlog run
\end{verbatim}
\end{framed}

System properties recognized by this analysis are:
\begin{itemize}
\item
\code{chord.inst.ctxt.kind}: the kind of context sensitivity to use for each
instance method (and all its locals).  It may be 'ci' (context insensitive), 'cs'
(k-CFA), or 'co' (k-object-sensitive).  It is 'ci' by default.
\item
\code{chord.stat.ctxt.kind}: the kind of context sensitivity to use for each
static method (and all its locals).  It may be 'ci' (context insensitive), 'cs'
(k-CFA), or 'co' (copy-context-sensitive).  It is 'ci' by default.
\item
\code{chord.ctxt.kind}: the kind of context sensitivity to use for each method
(and all its locals).  It may be one of 'ci', 'cs', or 'co', and serves as
shorthand for properties \code{chord.inst.ctxt.kind} and \code{chord.stat.ctxt.kind}.
\item
\code{chord.kobj.k} and \code{chord.kcfa.k}: the 'k' value to use for each object
allocation site and each method call site, respectively.
It is 1 by default.
\end{itemize}

The above command runs the most general form of context-sensitive
points-to/call-graph analysis,
allowing different kinds of context- and object-sensitivity to be mixed in the analysis of
the same program, but this flexibility comes at the cost of efficiency.
Hence, Chord provides three specialized forms of this analysis that more
efficiently handle certain common instantiations of this analysis that do not
arbitrarily mix different kinds of context- and object-sensitivity.  We next describe these
three specialized analyses.

If \code{chord.ctxt.kind} is 'ci' (the default value),
the above command essentially performs a purely context-insensitive
analysis; in this case, it is preferable to run the following more efficient
implementation of that analysis:

\begin{framed}
\begin{verbatim}
ant -Dchord.work.dir=<WORK_DIR> -Dchord.run.analyses=cspa-0cfa-dlog run
\end{verbatim}
\end{framed}

If \code{chord.ctxt.kind} is 'cs' and \code{chord.kcfa.k} is K,
the above command essentially performs a purely K-CFA analysis; in this case, it
is preferable to run the following more efficient implementation of that
analysis:

\begin{framed}
\begin{verbatim}
ant -Dchord.work.dir=<WORK_DIR> -Dchord.ctxt.kind=cs -Dchord.kcfa.k=K -Dchord.run.analyses=cspa-kcfa-dlog run
\end{verbatim}
\end{framed}

If \code{chord.ctxt.kind} is 'co' and \code{chord.kobj.k} is K,
the above command essentially performs a purely K-object-sensitive analysis; in
this case, it is preferable to run the following more efficient implementation
of that analysis:

\begin{framed}
\begin{verbatim}
ant -Dchord.work.dir=<WORK_DIR> -Dchord.ctxt.kind=co -Dchord.kobj.k=K -Dchord.run.analyses=cspa-kobj-dlog run
\end{verbatim}
\end{framed}

Each of the above four commands outputs the following relations:

\begin{itemize}
\item
{\it Context information:}
\begin{itemize}
\item
C is the domain of all abstract contexts and abstract objects.  Each
element in this domain is a sequence of zero or more sites, where each site can
be a call site or an object allocation site.  A sequence may have mixed call
sites and object allocation sites.  The most significant site (i.e. the first
site) in each sequence is called the head; the remaining sub-sequence is called
the tail.  The below three relations relate a sequence with its head and tail.
\item
CC subset (C $\times$ C) contains tuples \texttt{(c1,c2)} such that context
\texttt{c2} is the tail of context \texttt{c1}.
\item
CH subset (C $\times$ H) contains tuples \texttt{(c,h)} such that object
allocation site \texttt{h} is the head of context \texttt{c}.
\item
CI subset (C $\times$ I) contains tuples \texttt{(c,i)} such that call site
\texttt{i} is the head of context \texttt{c}.
\end{itemize}
\item
{\it Call-graph information:}
\begin{itemize}
\item
rootCM subset (C $\times$ M) contains tuples \texttt{(c,m)} such that method
\texttt{m} is an entry method in context \texttt{c}.
\item
reachableCM subset (C $\times$ M) contains tuples \texttt{(c,m)} such that
method \texttt{m} may be reachable in context \texttt{c}.
\item
CICM subset (C $\times$ I $\times$ C $\times$ M) contains tuples
\texttt{(c1,i,c2,m)} such that call site \texttt{i} in context \texttt{c1} may
call method \texttt{m2} in context \texttt{c2}.
\end{itemize}
\item
{\it Points-to information:}
\begin{itemize}
\item
FC subset (F $\times$ C) contains tuples \texttt{(f,o)} such that static field
(i.e. global variable) \texttt{f} may point to object \texttt{o}.
\item
CVC subset (C $\times$ V $\times$ C) contains tuples \texttt{(c,v,o)} such that
local variable \texttt{v} may point to object \texttt{o} in context \texttt{c}
of that variable's declaring method.  Note that both \texttt{o} and \texttt{c}
are elements of domain C.
\item
CFC subset (C $\times$ F $\times$ C) contains tuples \texttt{(o1,f,o2)} such
that instance field \texttt{f} of object \texttt{o1} may point to object
\texttt{o2}.  \end{itemize}
\end{itemize}

\subsection{Call Graph Variants}

Chord offers two variants of each of the context-insensitive and
context-sensitive baseline call graphs presented above, called {\it abbreviated
thread-insensitive} and {\it abbreviated thread-sensitive} call graphs.  We
explain these two variants for the context-insensitive case; they are analogous
for the context-sensitive case.  Consider the following example Java program:

\begin{framed}
\begin{verbatim}
public class T extends java.lang.Thread {
    static {
i1:     m1();
    }
    public static void m1() { }
    public static void m2() { }
    public static void m3() { }
    public void run() {
i2:     m2();
    }
    public static void main() {
i3:     m3();
        T t = ...;
i4:     t.start();
    }
}
\end{verbatim}
\end{framed}

The baseline call graph for this program is 
rootM = \{ main, \code{<clinit>} \}, 
reachableM = \{ main, \code{<clinit>}, m1, m2, m3, start, run \}, and
IM = \{ (i1, m1), (i2, m2), (i3, m3), (i4, start), (i5, run) \}, where methods
main, m1, m2, m3, and run are those declared in class C, method \code{<clinit>}
is the class initializer method of class C, and the
\code{java.lang.Thread.start()} method contains a call site labeled i5 that calls
\code{this.run()}.

An {\it abbreviated thread-insensitive} call graph excludes from the baseline call
graph all methods that are unreachable from the main method.  In other words, it
excludes all class initializer methods as well as methods that are reachable
only from such methods.
For the above example, this call graph is
rootM = \{ main \},
reachableM = \{ main, m2, m3, start, run \}, and
IM = \{ (i2, m2), (i3, m3), (i4, start), (i5, run) \}.  Notice that methods
\code{<clinit>} and m1 are omitted from reachableM, and call edge (i1, m1) is
omitted from IM, since m1 is unreachable from the main method.

An {\it abbreviated thread-sensitive} call graph excludes from the baseline call
graph all methods that are unreachable from the main method, like the
abbreviated thread-insensitive call graph, and additionally treats the
\code{java.lang.Thread.start()} method as a root method and does not include
call edges that target that method.
For the above example, this call graph is
rootM = \{ main, start \},
reachableM = \{ main, m2, m3, start, run \}, and
IM = \{ (i2, m2), (i3, m3), (i5, run) \}.  Notice that the start method is
added to rootM and call edge (i4, start) is omitted from IM.

The above two call graph variants are provided to simplify expressing
certain forms of reachability:
\begin{itemize}
\item
An {\it abbreviated} call graph allows to focus just on reachability
from the main method, as opposed to from class initializer methods as well;
this is used, for instance, in Chord's static datarace analysis to
suppress reporting races on statements reachable only from class initializer
methods, which are typically false races in practice.
\item
A {\it thread-sensitive} call graph additionally allows to focus just on
``intra-thread'' reachability; this is used, for instance, in Chord's static datarace
analysis to prevent reporting false races (t1,e1,t2,e2)
where statement e1 (or e2 resp.) is unreachable from abstract thread
t1 (or t2 resp.).
\end{itemize}

### 2011-05-19 main/doc/relation.tex
updated documentation

\xname{datalog}
\chapter{Datalog Analysis}
\label{chap:datalog}

A common way to rapidly prototype an analysis in Chord is using a declarative
logic-programming language called Datalog.  This chapter describes all aspects
of Datalog analyses in Chord.  Section \ref{sec:writing-datalog} explains how to
write and run a Datalog analysis.  Section \ref{sec:tuning-datalog} explains how
to tune its performance.  Finally, Section \ref{sec:bdd} explains the
representation of BDDs (Binary Decision Diagrams) which are a data structure
central to executing Datalog analyses.

\section{Writing a Datalog Analysis}
\label{sec:writing-datalog}

A Datalog analysis declares a bunch of input/output program relations, each over
one or more program domains, and provides a bunch of rules (constraints)
specifying how to compute the output relations from the input relations.  It can
be defined in any file with suffix {\tt .dlog} in any directory in the path
specified by property \code{chord.dlog.analysis.path}.  An example Datalog
analysis is shown below:

\texonly{\newpage}

\begin{framed}
{\small
\begin{verbatim}
    # name=datarace-dlog

    # Program domains
    .include "E.dom"
    .include "F.dom"
    .include "T.dom"

    # BDD variable order
    .bddvarorder E0xE1_T0_T1_F0

    # Input/intermediate/output program relations
    field(e:E0,f:F0) input
    write(e:E0) input
    reach(t:T0,e:E0) input
    alias(e1:E0,e2:E1) input
    escape(e:E0) input
    unguarded(t1:T0,e1:E0,t2:T1,e2:E1) input
    hasWrite(e1:E0,e2:E1)
    candidate(e1:E0,e2:E1) 
    datarace(t1:T0,e1:E0, t2:T1,e2:E1) output

    # Analysis constraints
    hasWrite(e1,e2) :- write(e1).
    hasWrite(e1,e2) :- write(e2).
    candidate(e1,e2) :- field(e1,f), field(e2,f), hasWrite(e1,e2), e1 <= e2.
    datarace(t1,e1,t2,e2) :- candidate(e1,e2), reach(t1,e1), reach(t2,e2), \
        alias(e1,e2), escape(e1), escape(e2), unguarded(t1,e1,t2,e2).
\end{verbatim}
}
\end{framed}

Any line that begins with a ``{\tt \#}" is regarded a comment, except a
line of the form ``\code{# name=<ANALYSIS_NAME>}'', which specifies the name
\code{<ANALYSIS_NAME>} of the Datalog analysis.
Each Datalog analysis is expected to have exactly one such line.
The above Datalog analysis is named {\tt datarace-dlog}.

The ``\code{.include "<DOM_NAME>.dom"}'' lines specify each domain named
\code{<DOM_NAME>} that is needed by the Datalog analysis, i.e., each domain over which
any relation that is input/output by the Datalog analysis is defined. 
The declaration of each such relation specifies the relation's schema:
all the domains over which the relation is defined.
If the same domain appears multiple times in a relation's schema then
contiguous integers starting from 0 must be used to distinguish them; for instance,
in the above example, {\tt candidate} is a binary relation, both of whose
attributes have domain {\tt E}, and they are distinguished as {\tt E0} and {\tt E1}.

Each relation is represented symbolically (as opposed to explicitly)
using a graph-based data structure called a Binary Decision Diagram (BDD for short).
Each domain containing N elements is assigned log2(N) BDD variables.
The size of a BDD and the efficiency of operations on it depends heavily
on the order of these BDD variables.
The ``\code{.bddvarorder <BDD_VAR_ORDER>}'' line in the Datalog analysis enables the Datalog
analysis writer to specify this order.
It must list all domains along with their numerical suffixes, separated
by `{\tt \_}' or `{\tt x}'.
Using a `{\tt \_}' between two domains, such as {\tt T0\_T1}, means that the BDD variables assigned
to domain {\tt T0} precede those assigned to domain {\tt T1} in the BDD variable
order for this Datalog analysis.
Using a `{\tt x}' between two domains, such as {\tt E0xE1}, means that the
BDD variables assigned to domains {\tt E0} and {\tt E1}
will be interleaved in the BDD variable order for this Datalog analysis.
See Section \ref{sec:bdd} for more details on BDD representations.

Each Datalog analysis rule is a Horn clause of the form
``{\tt R(t) :- R1(t1), ..., Rn(tn)}"
meaning that if relations {\tt R1}, ..., {\tt Rn} contain tuples {\tt t1}, ..., {\tt tn}
respectively, then relation {\tt R} contains tuple {\tt t}.
A backslash may be used at the end of a line to break long rules for readability.
The Datalog analysis solver bddbddb used in Chord does not apply any
sophisticated optimizations to simplify the rules; besides the BDD variable order,
the manner in which these rules are expressed heavily affects the performance of
the solver.  For instance, an important manual optimization involves breaking down
long rules into multiple shorter rules communicating via intermediate relations.
See Section \ref{sec:tuning-datalog} for hints on tuning the performance
of Datalog analyses.

Running a Datalog analysis is no different from running any other kind of analysis
in Chord.  See Chapter \ref{chap:running} for how to run an analysis.

\section{Tuning Performance}
\label{sec:tuning-datalog}

There are several tricks analysis writers can try to improve the
performance of bddbddb, the Datalog solver used by Chord, often
by several orders of magnitude.
Try these tricks by running the following command:

\begin{framed}
\begin{verbatim}
ant -Ddlog.file=<FILE> -Dwork.dir=<DIR> solve
\end{verbatim}
\end{framed}

where {\tt <FILE>} is the file defining the Datalog analysis
to be tuned, and {\tt <DIR>} is the directory containing the
program domains ({\tt *.dom} files) and program relations ({\tt *.bdd} files)
consumed by the analysis (this is by default the
\code{chord_output/bddbddb/} directory generated
by a previous run of Chord.

\begin{enumerate}

\item

Set properties \verb+noisy=yes+, \verb+tracesolve=yes+, and \verb+fulltracesolve=yes+
on the above command line and observe which rule gets ``stuck" (i.e., takes several seconds to solve).
\verb+fulltracesolve+ is seldom useful, but \verb+noisy+ and \verb+tracesolve+ are
often very useful.  Once you identify the rule that is getting stuck, it
will also tell you which relations and which domains used in that rule,
and which operation on them, is taking a long time to solve.  Then try
to fix the problem with that rule by doing either or both of the following:
\begin{itemize}
\item
Break down the rule into multiple rules by creating intermediate relations (the more
relations you have on the RHS of a rule the slower it generally takes to solve
that rule).
\item
Change the relative order of the domains of those
relations in the BDD variable order
(note that you can use either `\_' or `x' between a pair of domains).
\end{itemize}

\item

Once you have ensured that none of the rules is getting ``stuck",
you will notice that some rules are applied too many times, and so
although each application of the rule itself isn't taking too much
time, the cumulative time for the rule is too much.  After finishing
solving a Datalog analysis, bddbddb prints how long each rule took to
solve (both in terms of the number of times it was applied and the
cumulative time it took).  It sorts the rules in the order of the
cumulative time.  You need to focus on the rules that took the most
time to solve (they will be at the bottom of the list).  Assuming you
removed the problem of rules getting ``stuck", the rules will roughly
be in the order of the number of times they were applied.  Here is an
example:

\begin{framed}
\begin{verbatim}
OUT> Rule VH(u:V0,h:H0) :- VV(u:V0,v:V1), VH(v:V1,h:H0), VHfilter
(u:V0,h:H0).
OUT>    Updates: 2871
OUT>    Time: 6798 ms
OUT>    Longest Iteration: 0 (0 ms)
OUT> Rule IM(i:I0,m:M0) :- reachableI(i:I0), specIMV(i:I0,m:M0,v:V0), VH(v:V0,_:H0).
OUT>    Updates: 5031
OUT>    Time: 6972 ms
OUT>    Longest Iteration: 0 (0 ms)
\end{verbatim}
\end{framed}

Notice that the second rule was applied 5031 times whereas the first
was applied 2871 times.  More importantly, the second rule took 6972
milliseconds in all, compared to 6798 for the first rule.  Hence, you
should focus on the second rule first, and try to speed it up.  This
means that you should focus only on relations IM, reachableI, specIMV,
and VH, and the domains I0, M0, V0, and H0.  Any changes you make that
do not affect these relations and domains are unlikely to make your
solving faster.  In general, look at the last few rules, not just the
last one, and try to identify the ``sub-analysis" of the Datalog analysis
that seems problematic, and then focus on speeding up just that sub-
analysis.

\item

You can add the \verb+.split+ keyword at the end of certain rules as a
hint to bddbddb to decompose those rules into simpler ones that can be
solved faster.  You can also set property \verb+split_all_rules=yes+ as shorthand
for splitting all rules without adding the \verb+.split+ keyword to any of
them, though I seldom find splitting all rules helpful.

\item

You can try to decompose a single Datalog analysis file into two separate Datalog analysis
files.  Of course, you cannot separate mutually-recursive rules into two
different analyses, but if you unnecessarily club
together rules that could have gone into different analyses, then they
can put conflicting demands on bddbddb (e.g., on the BDD variable order).
So if rule 2 uses the result of rule 1 and rule 1 does not use the result of
rule 2, then put rule 1 and rule 2 in separate Datalog analyses.

\item

Observe the sizes of the BDDs representing the relations that are
input and output.  bddbddb prints both the number of tuples in each
relation and the number of nodes in the BDD.  Try changing the
BDD variable order for the domains of the relation, and observe how the
number of nodes in the BDD for that relation change.  You will
notice that some orders perform remarkably better than others.  Then
note down these orders as invariants that you will not violate as
you tweak other things.

\item

The relative order of values *within* domains (e.g.,
in domains named \verb+M+, \verb+H+, \verb+C+, etc. in Chord) affects the
performance of bddbddb, but
I've never tried changing this and studying its effect.  It might be
worth trying.  For instance, John Whaley's PLDI'04 paper describes a
specific way in which he numbers contexts (in domain \verb+C+) and that it was
fundamental to the speedup of his ``infinity"-CFA points-to analysis.

\item

Finally, it is worth emphasizing that BDDs are not magic.
If your algorithm itself is fundamentally hard to scale, then BDDs are
unlikely to help you a whole lot.  Secondly, many things are awkward to
encode as integers (e.g., the abstract contexts in domain \verb+C+ 
in Chord) or as Datalog rules.
For instance, I've noticed that summary-based context-sensitive program
analyses are hard to express in Datalog.  The may-happen-in-parallel
analysis provided in Chord shows a relatively simple kind of summary-based
analysis that uses the Reps-Horwitz-Sagiv tabulation algorithm.  But this
is as far as I could get---more complicated summary-based algorithms are
best written in Java itself instead of Datalog.
\end{enumerate}

\section{BDD Representation}
\label{sec:bdd}

>>>>

\xname{relation}
\chapter{Program Relation}
\label{chap:relation}

Finally, Section \ref{sec:bdd} explains the
representation of BDDs (Binary Decision Diagrams) which are a data structure
central to executing Datalog analyses.

\section{Constructing a Relation}

1. Using ProgramRel
2. Produced by analysis
3. Manual

\section{Operations on a Relation}

\section{BDD Representation Internals}
\label{sec:bdd}

### 2011-05-19 main/doc/user_guide.tex
updated documentation

@@

\setcounter{htmldepth}{2}
\setcounter{htmlautomenu}{3}

@@

\input{whatis_chord}
\input{getting_started}
\input{example}
\input{architecture}
\input{properties}
\input{predefined_analyses}
\input{analyses}
\input{dynamic_analysis}
\input{datalog_analysis}
\input{computing_scope}
\input{program_representation}
\input{acknowledgments}

>>>>

\newcommand\includeimage[2]{\texonly{\includegraphics[scale=#1]{#2}}\htmlonly{{\htmlimg{#2}}}}

@@

\newcommand\bsdlicense{\xlink{New BSD License}{http://www.opensource.org/licenses/bsd-license.php}}
\newcommand\chordweb{\url{http://jchord.googlecode.com/}}
\newcommand\buddy{\xlink{BuDDy}{http://buddy.sourceforge.net/}}
\newcommand\bddbddb{\xlink{bddbddb}{http://bddbddb.sourceforge.net/}}
\newcommand\ibmjvm{\xlink{IBM J9}{http://www.ibm.com/developerworks/java/jdk/}}
\newcommand\sunjvm{\xlink{Oracle HotSpot}{http://www.oracle.com/technetwork/java/javase/downloads/index.html}}
\newcommand\chordbinfile{\xlink{chord-bin-2.0.tar.gz}{http://jchord.googlecode.com/files/chord-bin-2.0.tar.gz}}
\newcommand\chordsrcfile{\xlink{chord-src-2.0.tar.gz}{http://jchord.googlecode.com/files/chord-src-2.0.tar.gz}}
\newcommand\chordlibfile{\xlink{chord-libsrc-2.0.tar.gz}{http://jchord.googlecode.com/files/chord-libsrc-2.0.tar.gz}}
\newcommand\ant{\xlink{Apache Ant}{http://ant.apache.org/}}
\newcommand\javassist{\xlink{Javassist}{http://www.javassist.org/}}
\newcommand\joeq{\xlink{Joeq}{http://joeq.sourceforge.net/}}

\setcounter{htmldepth}{3}
\setcounter{htmlautomenu}{4}

@@

%\texonly{\tableofcontents}

\xname{preface}
\part{Preface}

\input{organization}
\input{acks}

\xname{enduser_guide}
\part{Guide for End-Users}

\input{whatis}
\input{start}
\input{properties}
\input{setup}
\input{scope}
\input{predefined}

\xname{developer_guide}
\part{Guide for Developers}

\input{arch}
\input{program}
\input{project}
\input{writing}
\input{running}
\input{dynamic}
\input{datalog}
